### 2018-06-17 express2018/intro.tex
text

In the   \emph{weak} case (when  behavioural equivalences abstract from internal moves,
which practically is the most relevant case), however, 
Milner's proof method has severe syntactical limitations. 
To overcome such limitations, Sangiorgi proposes to replace
equations with  special inequations called
\emph{contractions} \cite{sangiorgi2015equations}. Contraction is a
preorder that, roughly, places some efficiency
constraints on processes.  Uniqueness of the solutions of a system of contractions
 is defined as with systems of equations:  
any two solutions must be bisimilar.
The difference with equations is in the meaning of solution:
in the case of contractions
the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem can be
removed.  One constraint that still remains in
\cite{sangiorgi2015equations} is on occurrences of the sum operator,
due to the failure of substitutivity of contraction w.r.t. such operator.

The main goal  of the work described in this paper is 
a rather  
 comprehensive formalisation  of the core of the theory of CCS 
 in the HOL
theorem prover (HOL4),  with a focus on the theory of unique solutions of contractions.
The formalisation however is not confined to the theory of  unique solutions, but embraces 
most of the 
core of the theory of CCS \cite{Mil89}
(partly because the theory of unique solutions
 relies on a number of more fundamental results):
indeed the formalisation encompasses the basic properties of strong and weak
bisimilarity (e.g. the fixed-point and substitutivity properties), 
their algebraic theory, various versions of ``bisimulation up to''
techniques (e.g., bisimulation up-to bisimilarity),
the main properties  of the rooted bisimilarity (the congruence induced by weak
bisimilarity, also called observation congruence). Concerning rooted bisimilarity, the formalisation
includes Hennessy and Deng lemmas, and two proofs that rooted bisimilarity is the largest
congruence included in bisimilarity:
one as in Milner's
book,  requiring the hypothesis that  no processes can use all labels; the other without
such hypothesis, essentially formalising van Glabbeek's paper \cite{vanGlabbeek:2005ur}
(such proof however follows the structure of the ordinal numbers, which cannot be handled
in HOL, and therefore it is restricted to finite-state processes ).
Similar theorems are proved for the rooted contraction preorder.

In this respect,  the work is an extensive experiment with the use of the HOL theorem prover and its
most recent developments, including a package  for expressing coinductive definitions.
The
work consists of about 20,000 lines of proof scripts in Standard ML.

From the CCS theory viewpoint, the formalisation has offered us the possibility of 
further refining the theory of unique solutions. 
In particular, the existing theory  placed limitations on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.  
We have thus refined the proof  technique based on contractions by moving to the 
\emph{rooted contraction}, that is, the coarsest (pre)congruence contained in the contraction
preorder.  Using rooted contraction one obtains a unique-solution theorem that is valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and without syntactic
constraints on the occurrences of sums.

@@

Another advantage of the formalisation is 
that we can take advantage of results about different 
equivalences or preorders that share a similar  proof structure. 
Examples are: the results that rooted bisimilarity and rooted contraction are,
respectively, the coarsest congruence contained in weak bisimilarity 
and the coarsest precongruence contained in the contraction  preorder; 
the result about unique solution of equations for weak bisimilarity that uses the
contraction preorder as an auxiliary relation, and other unique solution results (e.g., 
the one for rooted in which
the auxiliary relation is rooted contraction); various forms of enhancement of the bisimulation
proof method (the `up-to' techniques).  
In these cases, there are only a few places in which the HOL scripts have to be modified.
Then the successful termination of the scripts  gives us a guarantee that the proof is
completed,  removing the risk 
of overlooking or missing details as in hand-written proofs.

>>>>

In the   \emph{weak} case (when  behavioural equivalences abstract from internal moves,
which practically is the most relevant case), however, 
Milner's proof method has severe syntactical limitations. 
A way for  overcoming such limitations is to replace 
equations
 with  special inequations called
\emph{contractions} \cite{sangiorgi2015equations}. Contraction is a
preorder that, roughly, places some efficiency
constraints on processes.  Uniqueness of the solution of a system of contractions
 is defined as with systems of equations:  
any two solutions must be bisimilar.
The difference with equations is in the meaning of a solution:
in the case of contractions
the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem can be
removed.  One constraint that still remains in
\cite{sangiorgi2015equations} is on occurrences of the sum operator,
due to the failure of the substitutivity of contraction for  such an operator.

The main goal  of the work described in this paper is 
a rather  
 comprehensive formalisation  of the core of the theory of CCS 
 in the HOL
theorem prover (HOL4),  with a focus on the theory of unique solutions of contractions.
The formalisation however is not confined to the theory of  unique solutions, but embraces 
most of the 
core of the theory of CCS \cite{Mil89}
(partly because the theory of unique solutions
 relies on a number of more fundamental results):
indeed the formalisation encompasses the basic properties of strong and weak
bisimilarity (e.g. the fixed-point and substitutivity properties), 
their basic algebraic theory, various versions of ``bisimulation up to''
techniques (e.g., bisimulation up-to bisimilarity),
the main properties  of the rooted bisimilarity (the congruence induced by weak
bisimilarity, also called observation congruence), and of the expansion preorder (an
efficiency-like refinement of bisimilarity). Concerning rooted bisimilarity, the formalisation
includes Hennessy and Deng lemmas, and two proofs that rooted bisimilarity is the largest
congruence included in bisimilarity:
one as in Milner's
book,  requiring the hypothesis that  no processes can use all labels; the other without
such hypothesis, essentially formalising van Glabbeek's paper \cite{vanGlabbeek:2005ur}
(such a proof however follows the structure of the ordinal numbers, which cannot be handled
in HOL, and therefore it is restricted to finite-state processes ).
Similar theorems are proved for the rooted contraction preorder.
In this respect,  the work is an extensive experiment with the use of the HOL theorem prover and its
most recent developments, including a package  for expressing coinductive definitions.
The
work consists of about 20,000 lines of proof scripts in Standard ML.

From the CCS theory viewpoint, the formalisation has offered us the possibility of 
further refining the theory of unique solutions. 
In particular, the existing theory  placed limitations on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.  
We have thus refined the proof  technique based on contractions by moving to  
\emph{rooted contraction}, that is, the coarsest (pre)congruence contained in the contraction
preorder.  The resulting 
 unique-solution theorem is valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and places no 
constraints on the occurrences of sums.

@@

Another advantage of the formalisation is 
that we can take advantage of results about different 
equivalences or preorders that share a similar  proof structure. 
Examples are: the results that rooted bisimilarity and rooted contraction are,
respectively, the coarsest congruence contained in weak bisimilarity 
and the coarsest precongruence contained in the contraction  preorder; 
the result about unique solution of equations for weak bisimilarity that uses the
contraction preorder as an auxiliary relation, and other unique solution results (e.g., 
the one for rooted in which
the auxiliary relation is rooted contraction); various forms of enhancement of the bisimulation
proof method (the `up-to' techniques).  
In these cases,  moving between proofs there are only a few places in which the HOL scripts have to be modified.
Then the successful termination of the scripts  gives us a guarantee that the proof is
completed,  removing the risk 
of overlooking or missing details as in hand-written proofs.

### 2018-06-17 express2018/part2.tex
text

Now we show the formalized statements of Lemma~\ref{l:uptocon}, which
is quite long:\vspace{-1em}
\begin{alltt}
\begin{small}
\HOLTokenTurnstile{} (\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{E}. \HOLConst{WGS} \HOLBoundVar{E} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLBoundVar{E} \HOLFreeVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{Q} \HOLSymConst{\HOLTokenContracts{}} \HOLBoundVar{E} \HOLFreeVar{Q}) \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{C}.
       \HOLConst{GCONTEXT} \HOLBoundVar{C} \HOLSymConst{\HOLTokenImp{}}
       (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{l} \HOLBoundVar{R}.
            \HOLBoundVar{C} \HOLFreeVar{P} \HOLTokenWeakTransBegin\HOLConst{label} \HOLBoundVar{l}\HOLTokenWeakTransEnd \HOLBoundVar{R} \HOLSymConst{\HOLTokenImp{}}
            \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{C\sp{\prime}}.
                \HOLConst{GCONTEXT} \HOLBoundVar{C\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{R} \HOLSymConst{\HOLTokenContracts{}} \HOLBoundVar{C\sp{\prime}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenConj{}}
                (\HOLConst{WEAK_EQUIV} \HOLSymConst{\HOLTokenRCompose{}} (\HOLTokenLambda{}\HOLBoundVar{x} \HOLBoundVar{y}. \HOLBoundVar{x} \HOLTokenWeakTransBegin\HOLConst{label} \HOLBoundVar{l}\HOLTokenWeakTransEnd \HOLBoundVar{y})) (\HOLBoundVar{C} \HOLFreeVar{Q})
                  (\HOLBoundVar{C\sp{\prime}} \HOLFreeVar{Q})) \HOLSymConst{\HOLTokenConj{}}
       \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R}.
           \HOLBoundVar{C} \HOLFreeVar{P} \HOLTokenWeakTransBegin\HOLSymConst{\ensuremath{\tau}}\HOLTokenWeakTransEnd \HOLBoundVar{R} \HOLSymConst{\HOLTokenImp{}}
           \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{C\sp{\prime}}.
               \HOLConst{GCONTEXT} \HOLBoundVar{C\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{R} \HOLSymConst{\HOLTokenContracts{}} \HOLBoundVar{C\sp{\prime}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenConj{}}
               (\HOLConst{WEAK_EQUIV} \HOLSymConst{\HOLTokenRCompose{}} \HOLConst{EPS}) (\HOLBoundVar{C} \HOLFreeVar{Q}) (\HOLBoundVar{C\sp{\prime}} \HOLFreeVar{Q})
\end{small}
\end{alltt}
\vspace{-1em} Traces were involved in the proof of above lemma by using
the following ``unfolding'' lemma:\vspace{-1em}
\begin{alltt}
\begin{small}
\HOLTokenTurnstile{} \HOLConst{GCONTEXT} \HOLFreeVar{C} \HOLSymConst{\HOLTokenConj{}} \HOLConst{WGS} \HOLFreeVar{E} \HOLSymConst{\HOLTokenConj{}} \HOLConst{TRACE} ((\HOLFreeVar{C} \HOLSymConst{\HOLTokenCompose} \HOLConst{FUNPOW} \HOLFreeVar{E} \HOLFreeVar{n}) \HOLFreeVar{P}) \HOLFreeVar{xs} \HOLFreeVar{P\sp{\prime}} \HOLSymConst{\HOLTokenConj{}}
   \HOLConst{LENGTH} \HOLFreeVar{xs} \HOLSymConst{\HOLTokenLeq{}} \HOLFreeVar{n} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{C\sp{\prime}}.
       \HOLConst{GCONTEXT} \HOLBoundVar{C\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} (\HOLFreeVar{P\sp{\prime}} \HOLSymConst{=} \HOLBoundVar{C\sp{\prime}} \HOLFreeVar{P}) \HOLSymConst{\HOLTokenConj{}}
       \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Q}. \HOLConst{TRACE} ((\HOLFreeVar{C} \HOLSymConst{\HOLTokenCompose} \HOLConst{FUNPOW} \HOLFreeVar{E} \HOLFreeVar{n}) \HOLBoundVar{Q}) \HOLFreeVar{xs} (\HOLBoundVar{C\sp{\prime}} \HOLBoundVar{Q})
\end{small}
\end{alltt}
It roughly says, for any context $C$ and weakly guarded expression
$E$, if $C \circ E^n[P] \overset{xs}{\Longrightarrow} P'$ and the length
of actions $xs \leqslant n$, then $P$ has the form of $C'[P]$, meaning
it's not touched during the transitions. Without traces it is
impossible to reason on the length of intermediate actions behind weak transitions.

With all above work, finally we can formally prove
Theorem~\ref{t:contraBisimulationU}: (here \texttt{WGS} stands for
`weakly guarded context with weakly-guarded sums')
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{WGS} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{Q}
\end{alltt}

@@

The formal proof of ``unique solution of rooted contractions theorem''
(Theorem~\ref{t:rcontraBisimulationU}) has the
same initial proof steps as Theorem~\ref{t:contraBisimulationU}; 
it then requires a
few steps to handle  rooted bisimilarity in the conclusion. 
At the end the
two proofs are very similar, mostly because the only property we need
from (rooted) contraction is its precongruence. 
(Having proved
the precongruence property for the rooted contraction, 
we can use the
plain version of weakly-guarded expressions with sums
included.) Below is the formally verified version of
Theorem~\ref{t:rcontraBisimulationU}:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{WG} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenObsContracts} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenObsContracts} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{Q}
\end{alltt}

Having no constraints on sums, the result is
 similar with Milner's original `unique solution of
equations' theorem for \emph{strong} bisimilarity ($\sim$)~--- 
the same
weakly guarded  condition (\texttt{WG}) is required:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{WG} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenStrongEQ} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenStrongEQ} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenStrongEQ} \HOLBoundVar{Q}
\end{alltt}

>>>>

We show the  formalized statement of Lemma~\ref{l:uptocon}:\vspace{-1em}
\begin{alltt}
\begin{small}
\HOLTokenTurnstile{} (\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{E}. \HOLConst{WGS} \HOLBoundVar{E} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLBoundVar{E} \HOLFreeVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{Q} \HOLSymConst{\HOLTokenContracts{}} \HOLBoundVar{E} \HOLFreeVar{Q}) \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{C}.
       \HOLConst{GCONTEXT} \HOLBoundVar{C} \HOLSymConst{\HOLTokenImp{}}
       (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{l} \HOLBoundVar{R}.
            \HOLBoundVar{C} \HOLFreeVar{P} \HOLTokenWeakTransBegin\HOLConst{label} \HOLBoundVar{l}\HOLTokenWeakTransEnd \HOLBoundVar{R} \HOLSymConst{\HOLTokenImp{}}
            \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{C\sp{\prime}}.
                \HOLConst{GCONTEXT} \HOLBoundVar{C\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{R} \HOLSymConst{\HOLTokenContracts{}} \HOLBoundVar{C\sp{\prime}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenConj{}}
                (\HOLConst{WEAK_EQUIV} \HOLSymConst{\HOLTokenRCompose{}} (\HOLTokenLambda{}\HOLBoundVar{x} \HOLBoundVar{y}. \HOLBoundVar{x} \HOLTokenWeakTransBegin\HOLConst{label} \HOLBoundVar{l}\HOLTokenWeakTransEnd \HOLBoundVar{y})) (\HOLBoundVar{C} \HOLFreeVar{Q})
                  (\HOLBoundVar{C\sp{\prime}} \HOLFreeVar{Q})) \HOLSymConst{\HOLTokenConj{}}
       \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R}.
           \HOLBoundVar{C} \HOLFreeVar{P} \HOLTokenWeakTransBegin\HOLSymConst{\ensuremath{\tau}}\HOLTokenWeakTransEnd \HOLBoundVar{R} \HOLSymConst{\HOLTokenImp{}}
           \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{C\sp{\prime}}.
               \HOLConst{GCONTEXT} \HOLBoundVar{C\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{R} \HOLSymConst{\HOLTokenContracts{}} \HOLBoundVar{C\sp{\prime}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenConj{}}
               (\HOLConst{WEAK_EQUIV} \HOLSymConst{\HOLTokenRCompose{}} \HOLConst{EPS}) (\HOLBoundVar{C} \HOLFreeVar{Q}) (\HOLBoundVar{C\sp{\prime}} \HOLFreeVar{Q})
\end{small}
\end{alltt}
\vspace{-1em} Traces are introduced in the proof of above lemma via 
the following `unfolding' lemma:\vspace{-1em}
\begin{alltt}
\begin{small}
\HOLTokenTurnstile{} \HOLConst{GCONTEXT} \HOLFreeVar{C} \HOLSymConst{\HOLTokenConj{}} \HOLConst{WGS} \HOLFreeVar{E} \HOLSymConst{\HOLTokenConj{}} \HOLConst{TRACE} ((\HOLFreeVar{C} \HOLSymConst{\HOLTokenCompose} \HOLConst{FUNPOW} \HOLFreeVar{E} \HOLFreeVar{n}) \HOLFreeVar{P}) \HOLFreeVar{xs} \HOLFreeVar{P\sp{\prime}} \HOLSymConst{\HOLTokenConj{}}
   \HOLConst{LENGTH} \HOLFreeVar{xs} \HOLSymConst{\HOLTokenLeq{}} \HOLFreeVar{n} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{C\sp{\prime}}.
       \HOLConst{GCONTEXT} \HOLBoundVar{C\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} (\HOLFreeVar{P\sp{\prime}} \HOLSymConst{=} \HOLBoundVar{C\sp{\prime}} \HOLFreeVar{P}) \HOLSymConst{\HOLTokenConj{}}
       \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Q}. \HOLConst{TRACE} ((\HOLFreeVar{C} \HOLSymConst{\HOLTokenCompose} \HOLConst{FUNPOW} \HOLFreeVar{E} \HOLFreeVar{n}) \HOLBoundVar{Q}) \HOLFreeVar{xs} (\HOLBoundVar{C\sp{\prime}} \HOLBoundVar{Q})
\end{small}
\end{alltt}
It roughly says that, for any context $C$ and weakly guarded expression
$E$, if $C [\, E^n[P]\,] \overset{xs}{\Longrightarrow} P'$ and the length
of actions $xs \leqslant n$, then $P$ has the form of $C'[P]$ (meaning
that $P$ is not touched during the transitions). Traces are necessary
 to  reason about the length of intermediate actions in weak transitions.

With all above work, finally we can formally prove
Theorem~\ref{t:contraBisimulationU} (here \texttt{WGS} stands for
`weakly guarded context with weakly-guarded sums'):
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{WGS} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{Q}
\end{alltt}

@@

The formal proof of ``unique solution of rooted contractions theorem''
(Theorem~\ref{t:rcontraBisimulationU}) has the
same initial proof steps as Theorem~\ref{t:contraBisimulationU}; 
it then requires a
few more steps to handle  rooted bisimilarity in the conclusion. 
Overall  the
two proofs are very similar, mostly because the only property we need
from (rooted) contraction is its precongruence. 
 Below is the formally verified version of
Theorem~\ref{t:rcontraBisimulationU}
(having proved
the precongruence of rooted contraction, 
we can use  weakly-guarded expressions,   without constraints on  sums;
that is, \texttt{WG} in place of \texttt{WGS}):
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{WG} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenObsContracts} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenObsContracts} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{Q}
\end{alltt}
Having removed the  constraints on sums, the result is
 similar to Milner's original `unique solution of
equations' theorem for \emph{strong} bisimilarity ($\sim$)~--- 
the same
weakly guarded  condition (\texttt{WG}) is required:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{WG} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenStrongEQ} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenStrongEQ} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenStrongEQ} \HOLBoundVar{Q}
\end{alltt}

### 2018-06-17 express2018/related.tex
text

Monica Nesi was the first in this area, she formalized both pure CCS \cite{Nesi:1992ve} and Value-Passing CCS
\cite{Nesi:2017wo} in old versions of HOL theorem prover.
%\footnote{Part of
% the proof scripts for pure CCS now is available at \url{https://github.com/binghe/informatica-public/tree/master/CCS2/CCS-Nesi}.}.
Her main focus was on
 decision procedures (as ML functions) for
automatically proving bisimilarities between CCS
processes. We
first ported her work pure CCS formalisation to HOL4 with essential
changes and new contents.

Bengtson,  Parrow and Weber
have made a substantial formalisation work 
on CCS, $\pi$-calculi
\cite{bengtson2010formalising,bengtson2007completeness,bengtson2007formalising}
and $\psi$-calculi 
using Isabelle/HOL and its nominal datatype package, with main focus on the handling of name binders.
% Jesper Bengtson's PhD thesis 
%  \cite{bengtson2010formalising} looks  at 
% formalizations of CCS, $\pi$-calculi and $\psi$-calculi
% into Isabelle/HOL and its
% nominal logic. For CCS, 
% he has 
%  formalized
%  basic properties for strong/weak equivalence (congruence, basic
%  algebraic laws), for a  and observational
% congruence, however the CCS syntax doesn't have constants
% (or recursion operator), using replication in place of recursion.
% The formalisation effort has then been continued 
%
% On the other side, Jesper
% Bengtson and Joachim Parrow made great progress on $\pi$-calculus
% formalization and 
% proved that the algebraic axiomatization of bisimulation
% equivalence in the $\pi$-calculi is sound and
% complete. \cite{bengtson2007completeness} 
Other formalisations in this area include the work of Solange
Coupet-Grimal\footnote{\url{https://github.com/coq-contribs/ccs}} in Coq
and Chaudhuri et al.\;\cite{chaudhuri2015lightweight} in Abella. The
latter focuses on `bisimulation up-to' techniques for strong bisimilarity 
for CCS and $\pi$-calculus.
% Formalisations less related to ours
% include Kahsai and Miculan \cite{kahsai2008implementing} for the spi
% calculus  in Isabelle, and Hirschkoff \cite{hirschkoff1997full} for the $\pi$-calculus in Coq.

>>>>

The closest of our work is Monica Nesi's \cite{Nesi:1992ve,Nesi:2017wo}. She has made
the first formalization of CCS,  using early versions of the HOL theorem prover.
% both pure CCS  and Value-Passing CCS
%\cite{Nesi:2017wo} in old versions of HOL theorem prover.
%\footnote{Part of
% the proof scripts for pure CCS now is available at \url{https://github.com/binghe/informatica-public/tree/master/CCS2/CCS-Nesi}.}. 
Her main focus is  on
 decision procedures (as ML functions) for
automatically proving bisimilarities between CCS
processes.  Her work has been a basis for ours. However, the differences are substantial,
both because we have used the most recent version of HOL, and because of the spectrum of
the theory considered.

Bengtson,  Parrow and Weber
have made a substantial formalisation work 
on CCS, $\pi$-calculi
and $\psi$-calculi 
using Isabelle/HOL and its nominal datatype package, with main focus on the handling of
name binders \cite{bengtson2010formalising,bengtson2007completeness,bengtson2007formalising}.
% Jesper Bengtson's PhD thesis 
%  \cite{bengtson2010formalising} looks  at 
% formalizations of CCS, $\pi$-calculi and $\psi$-calculi
% into Isabelle/HOL and its
% nominal logic. For CCS, 
% he has 
%  formalized
%  basic properties for strong/weak equivalence (congruence, basic
%  algebraic laws), for a  and observational
% congruence, however the CCS syntax doesn't have constants
% (or recursion operator), using replication in place of recursion.
% The formalisation effort has then been continued 
%
% On the other side, Jesper
% Bengtson and Joachim Parrow made great progress on $\pi$-calculus
% formalization and 
% proved that the algebraic axiomatization of bisimulation
% equivalence in the $\pi$-calculi is sound and
% complete. \cite{bengtson2007completeness} 
Other formalisations in this area include the work of Solange
Coupet-Grimal\footnote{\url{https://github.com/coq-contribs/ccs}} in Coq
and Chaudhuri et al.\;\cite{chaudhuri2015lightweight} in Abella. The
latter focuses on `bisimulation up-to' techniques for strong bisimilarity 
for CCS and $\pi$-calculus.
% Formalisations less related to ours
% include Kahsai and Miculan \cite{kahsai2008implementing} for the spi
% calculus  in Isabelle, and Hirschkoff \cite{hirschkoff1997full} for the $\pi$-calculus in Coq.
